### 2015-05-10 swmpr_manu.Rnw
removed all footnotes

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS})\footnote{\url{http://www.nerrs.noaa.gov/}}.  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP}\footnote{\url{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \ac{SWMP} data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal\footnote{\url{http://cdmo.baruch.sc.edu/}}, where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data management.  As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.    Prior to any data request to the \ac{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the \ac{CDMO} website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with \ac{CDMO} before using the data retrieval functions in SWMPr.  Web services\footnote{\url{http://cdmo.baruch.sc.edu/webservices.cfm}} are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

@@

The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be downloaded from the \ac{CRAN}\footnote{\url{http://cran.r-project.org/web/packages/SWMPr/index.html} for the \ac{CRAN} release, the development version can be installed  from \url{https://github.com/fawda123/SWMPr}.} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

@@

Two approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services. To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.\footnote{Follow instructions here: \url{http://cdmo.baruch.sc.edu/webservices.cfm}}  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.

@@

The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.  This approach is most appropriate for large, specific data requests. The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

@@

A sample dataset can be downloaded for use with the examples below.\footnote{\url{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}}  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.  Processed verions of the raw data are also included with the package as binary data files (RData).  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags.\footnote{\ac{CDMO} flags: \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}} Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.  The \ac{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \ac{QAQC} flag (i.e., passed initial \ac{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages\footnote{See the listing on \ac{CRAN} for package imports and dependencies for SWMPr: \url{http://cran.r-project.org/web/packages/SWMPr/index.html}}.  These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

@@

The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.  The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).  An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.  The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.  For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).  The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).  For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.  A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (e.g., 1 if the the first observation is at sunrise for a diurnal cycle).\footnote{The value is passed to the \texttt{ts} function, \url{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}}  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

@@

Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \ac{CDMO}.  All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files.\footnote{Data processing script: \url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}} This resulted in a single \texttt{swmpr} object for each parameter at each site.  All files were then uploaded to a remote server for online access.  An additional R script\footnote{Metabolism processing script: \url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}} was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.  Two water quality sites with the longest time series at each reserve were used.  Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.  All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods. Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.  More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.

@@

The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data.  The first web application\footnote{Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}} evaluates trends in \ac{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}).  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.  The trend direction is blue for decreasing and red for increasing with time.  The second application\footnote{Summarize trends in \ac{SWMP} parameters for an individual site: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary/}} provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function (output is identical to \cref{fig:summary_ex}). Drop down menus can be used to select the station, date range, and parameter for plotting. The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.  Both of the apps provide previously unavailable utilities to interact with \ac{SWMP} data using an intuitive online interface.  This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package. The applications have been used extensively with over 300 hours noted in a single month.

@@

@@

All \ac{SWMP} data from 1995 through 2014, one file per site. See the help file for \texttt{import\_remote} for more information: \href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/}

@@

Source code as an R script for reproducing figures and tables in the manuscript: \href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R}

@@

Data used to create the figures and tables.  Files are in the .RData format: \href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data}

>>>>

% Remove comment for double spacing
\usepackage{setspace} 
\doublespacing

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \ac{SWMP} data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data management.  As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.    Prior to any data request to the \ac{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the \ac{CDMO} website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with \ac{CDMO} before using the data retrieval functions in SWMPr.  Web services are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

@@

The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be downloaded from the \ac{CRAN} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

@@

Two approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services. To do so, the IP address for the computer making the request must be registered by following instructions on the \ac{CDMO} website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.

@@

The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.  This approach is most appropriate for large, specific data requests. The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \ac{CDMO} website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

@@

A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.  The \ac{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \ac{QAQC} flag (i.e., passed initial \ac{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}). These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

@@

The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.  The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).  An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.  The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.  For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).  The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).  For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.  A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.  For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\texttt{ts}} function for details).  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

@@

Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \ac{CDMO}.  All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files (see \nameref{met_scripts}). This resulted in a single \texttt{swmpr} object for each parameter at each site.  All files were then uploaded to a remote server for online access.  An additional R script (see \nameref{met_scripts}) was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.  Two water quality sites with the longest time series at each reserve were used.  Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.  All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods. Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.  More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.

@@

The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data.  The first web application evaluates trends in \ac{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.  The trend direction is blue for decreasing and red for increasing with time.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to \cref{fig:summary_ex}. Drop down menus can be used to select the station, date range, and parameter for plotting. The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.  Both of the apps provide previously unavailable utilities to interact with \ac{SWMP} data using an intuitive online interface.  This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package. The applications have been used extensively with over 300 hours noted in a single month.

@@

\subsection*{Sample data files}
\label{supp_zip}

\href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{https://s3.amazonaws.com/swmpexdata/zip\_ex.zip} A compressed data folder is provided to show the format provided by the zip downloads feature of the \ac{CDMO}.  These data can be retrieved for use in R with the \texttt{import\_local} function.  The same data are included with the SWMPr package as binary RData files.

@@

\href{https://s3.amazonaws.com/swmpalldata/}{https://s3.amazonaws.com/swmpalldata/} All \ac{SWMP} data from 1995 through 2014, one file per site. See the help file for \texttt{import\_remote} for more information.

\subsection*{Metabolism scripts}
\label{met_scripts}

\url{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R} Data processing script for all data requested from the CDMO using the zip downloads feature.
\url{https://gist.github.com/fawda123/4fc51c2cb86341ed9291} Metabolism script for all data from the previous script.

@@

\href{https://github.com/fawda123/swmpr_manu/blob/master/swmpr_manu_code.R}{https://github.com/fawda123/swmpr\_manu/blob/master/swmpr\_manu\_code.R} Source code as an R script for reproducing figures and tables in the manuscript.

@@

\href{https://github.com/fawda123/swmpr_manu/tree/master/data}{https://github.com/fawda123/swmpr\_manu/tree/master/data} Data used to create the figures and tables.  Files are in the .RData format.
